---
title: "Assignment 4: Data Wrangling"
author: "Reed Leon-Hinton"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, Feb 16 @ 11:59pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1 Setting up the working directory and loading the required packages
getwd()

# clearing the environment (It's a pet peeve)
remove(list = ls())

# install.packages("tidyverse")
library(tidyverse)
# install.packages("lubridate")
library(lubridate)

# importing the datasets
o3_2018 <- read.csv(file = "./Data/Raw/EPAair_O3_NC2018_raw.csv", 
                    stringsAsFactors = TRUE)
o3_2019 <- read.csv(file = "./Data/Raw/EPAair_O3_NC2019_raw.csv", 
                    stringsAsFactors = TRUE)
pm25_2018 <- read.csv(file = "./Data/Raw/EPAair_PM25_NC2018_raw.csv", 
                      stringsAsFactors = TRUE)
pm25_2019 <- read.csv(file = "./Data/Raw/EPAair_PM25_NC2019_raw.csv", 
                      stringsAsFactors = TRUE)

#2 looking at the details of the datasets

# looking at the O3 2018 dataset
dim_o3_2018 <- dim(o3_2018)
colnames(o3_2018)
summary(o3_2018)
lapply(o3_2018, class)

# looking at the O3 2019 dataset
dim_o3_2019 <- dim(o3_2019)
colnames(o3_2019)
summary(o3_2019)
lapply(o3_2019, class)


# looking at the PM 2.5 2018 dataset
dim_pm25_2018 <- dim(pm25_2018)
colnames(pm25_2018)
summary(pm25_2018)
lapply(pm25_2018, class)


# looking at the PM 2.5 2019 dataset
dim_pm25_2019 <- dim(pm25_2019)
colnames(pm25_2019)
summary(pm25_2019)
lapply(pm25_2019, class)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 changing the format of the date column in each dataset
o3_2018$Date <- mdy(o3_2018$Date)
o3_2019$Date <- mdy(o3_2019$Date)
pm25_2018$Date <- mdy(pm25_2018$Date)
pm25_2019$Date <- mdy(pm25_2019$Date)

#4 creating processed datasets with the selected columns only
o3_2018_p <- select(.data = o3_2018,
                    Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                    COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
o3_2019_p <- select(.data = o3_2019,
                    Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                    COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
pm25_2018_p <- select(.data = pm25_2018,
                      Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                      COUNTY, SITE_LATITUDE, SITE_LONGITUDE)
pm25_2019_p <- select(.data = pm25_2019,
                      Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                      COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5 setting the value of the AQS_PARAMETER_DESC column to PM2.5
pm25_2018_p$AQS_PARAMETER_DESC <- as.factor("PM2.5")
pm25_2019_p$AQS_PARAMETER_DESC <- as.factor("PM2.5")

#6 creating the processed datasets in the processed folder
write.csv(o3_2018_p, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(o3_2019_p, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(pm25_2018_p, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(pm25_2019_p, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7 combining the datasets with the rbind function
o3_pm25_p <- rbind(o3_2018_p, o3_2019_p, pm25_2018_p, pm25_2019_p)

#8 completing the pipe function to meet the conditions specified
o3_pm25_p <- o3_pm25_p %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
                          "Hattie Avenue", "Clemmons Middle", "Mendenhall School", 
                          "Frying Pan Mountain", "West Johnston Co.", 
                          "Garinger High School", "Castle Hayne", "Pitt Agri. Center", 
                          "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise_at(vars(DAILY_AQI_VALUE, SITE_LATITUDE, SITE_LONGITUDE), list(name = mean)) %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  mutate(Year = year(Date)) %>%
  ungroup() %>%
  droplevels()
dim(o3_pm25_p) #just checking to see if I got the 14,752 x 9.

#9 spreading the dataset
o3_pm25_p <- pivot_wider(o3_pm25_p, names_from = AQS_PARAMETER_DESC, 
                         values_from = DAILY_AQI_VALUE_name)

colnames(o3_pm25_p)[4] <- "SITE_LATITUDE" # fixing the unusual naming resulting from the 
                                          # process where I meaned the values

colnames(o3_pm25_p)[5] <- "SITE_LONGITUDE" # doing the same as above

#10 looking at the dimensions of the dataset
final_dimensions <- dim(o3_pm25_p)
print(paste("The final dataset has", final_dimensions[1], "recorded values for", 
            final_dimensions[2], "variables."))

#11
write.csv(o3_pm25_p, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12
o3_pm25_p_summaries <- 
  o3_pm25_p %>%
  group_by(Site.Name, Month, Year) %>%
  summarise_at(vars(PM2.5, Ozone), list(name = mean)) %>%
  drop_na(Month, Year) %>%
  ungroup()

# fixing the naming convention like I did previously in the assignment
colnames(o3_pm25_p)[4] <- "PM2.5"
colnames(o3_pm25_p)[5] <- "Ozone"

#13 
summary_dimensions <- dim(o3_pm25_p_summaries)
print(paste("The summary dataset has", summary_dimensions[1], "recorded values for", 
            summary_dimensions[2], "variables."))


```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: na.omit drops all values which have NA associated with every row, whereas drop_na only drops rows with NA values from the columns specified. We get many 207 more observations when we use drop_na rather than na.omit. drop_na works better in this situation because we want to keep the NA values for Ozone and PM2.5, but not for the month and year. 
